PineAPPL:\\ combining EW and QCD corrections for PDF fits

### 2020-08-24 paper/sec-pineappl.tex
Clarify with PineAPPL does a bit

In experiments, but also for many calculations where the phase-space integration is performed using MC techniques, finite statistics does not allow us to reconstruct the exact dependence of the cross section on the observable $\mathcal{O}$.
Instead, it it sufficient to approximate the derivative using a piecewise-constant function,
\begin{equation}
W_{ab}^{(k,l,m,n)} \left( x_1, x_2, Q^2, \mathcal{O} \right) \approx \sum_{o=1}^M \frac{\Theta (\mathcal{O}_o^\mathrm{min} \le \mathcal{O} < \mathcal{O}_o^\mathrm{max})}{\mathcal{O}_o^\mathrm{max} - \mathcal{O}_o^\mathrm{min}} w_{ab}^{(k,l,m,n,o)} \left( x_1, x_2, Q^2 \right) \text{,}
\end{equation}
which uses $M$ bins with limits $\{ \mathcal{O}_o^\mathrm{min}, \mathcal{O}_o^\mathrm{max} \}_{o=1}^M$ to partition a finite range of the observable,
\begin{equation}
\mathcal{O}_0^\mathrm{min} < \mathcal{O}_0^\mathrm{max} = \mathcal{O}_1^\mathrm{min} < \ldots < \mathcal{O}_{M-1}^\mathrm{max} = \mathcal{O}_M^\mathrm{min} < \mathcal{O}_M^\mathrm{max} \text{.}
\label{eq:bins-of-diff-xsection}
\end{equation}
The Ellis--Sexton scale $Q^2$, if chosen dynamically, depends on the phase space, however we assume the fractions $\xi_\mathrm{R}$ and $\xi_\mathrm{F}$ to be  phase-space constants in any case.
This allows for variations around the central scale choice $\xi_\mathrm{R} = \xi_\mathrm{F} = 1$, but it does not otherwise allow for arbitrary changes of the scale.
The terms with powers $m > 0$ and $n > 0$ vanish for the central scale choice and are only required for variations of the factorisation and renormalisation scales.
To estimate the perturbative QCD uncertainty --- no EW uncertainty is covered by this method --- one typically uses scale variations. Common prescriptions are
7-point and 9-points scale variations, which evaluate the cross section using
respectively the following values
  \begin{align}
    (\xi_\mathrm{R}, \xi_\mathrm{F})_{\mathrm{7-pt}}
    & \in
    \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr) \right\} \text{;}
    \label{eq:7pt}\\
    (\xi_\mathrm{R}, \xi_\mathrm{F})_{\mathrm{9-pt}}
    & \in
    \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr), \bigl( \tfrac{1}{2}, 2 \bigr), \bigl(2, \tfrac{1}{2} \bigr) \right\}  \text{.}
    \label{eq:9pt}
   \end{align}
The (asymmetric) uncertainties are then given as the minimum and maximum value (the envelope), measured from the central value $(1, 1)$.
As is clear from eq.~\eqref{eq:expansion}, the EW coupling $\alpha$ is assumed not to be a dynamically varying coupling, but instead a constant over phase space.
This, however, includes the most common choices of the coupling, which are (not necessarily in this order), $\alpha (0)$, $\alpha (M_\mathrm{Z})$, and $\alpha_{G_\mu}$.

The problem that \textsc{PineAPPL} solves can now be described: approximately reconstruct the functions
\begin{equation}
w_{ab}^{(k,l,m,n,o)} \left( x_1, x_2, Q^2 \right)
\label{eq:weight-map}
\end{equation}
from a set of $N$ function evaluations for specific momentum fractions, scales, and values of the observable
\begin{equation}
\left\{ x_1^{(i)}, x_2^{(i)}, Q^2_i, \mathcal{O}_i \right\}_{i=1}^N \text{,} \label{eq:phase-space-weights}
\end{equation}
given by the MC integrator.
Using eq.~\eqref{eq:expansion} and
\begin{multline}
\frac{\mathrm{d} \sigma}{\mathrm{d} \mathcal{O}} (\mathcal{O}, \xi_\mathrm{R}, \xi_\mathrm{F}) \\
= \sum_{a,b} \int_0^1 \mathrm{d} x_1 \int_0^1 \mathrm{d} x_2 \int_{Q^2_\mathrm{min}}^{Q^2_\mathrm{max}} \mathrm{d} Q^2 \, f_a (x_1, \xi_\mathrm{F}^2 Q^2) f_b (x_2, \xi_\mathrm{F}^2 Q^2) \sigma_{ab} (x_1, x_2, Q^2, \xi_\mathrm{R}, \xi_\mathrm{F}) \text{,}
\label{eq:pineappl-convolution}
\end{multline}
\textsc{PineAPPL} can then quickly calculate hadronic cross sections for
an arbitrary number of PDF sets and perform scale variations.

>>>>

In experiments, but also for many calculations where the phase-space integration is performed using MC techniques, finite statistics does not allow us to reconstruct the exact dependence of the cross section on the observable $\mathcal{O}$.
Instead, it it sufficient to approximate the derivative using a piecewise-constant function,
\begin{equation}
W_{ab}^{(k,l,m,n)} \left( x_1, x_2, Q^2, \mathcal{O} \right) \approx \sum_{o=1}^M \frac{\Theta (\mathcal{O}_o^\mathrm{min} \le \mathcal{O} < \mathcal{O}_o^\mathrm{max})}{\mathcal{O}_o^\mathrm{max} - \mathcal{O}_o^\mathrm{min}} w_{ab}^{(k,l,m,n,o)} \left( x_1, x_2, Q^2 \right) \text{,}
\label{eq:differential-cross-section}
\end{equation}
which uses $M$ bins with limits $\{ \mathcal{O}_o^\mathrm{min}, \mathcal{O}_o^\mathrm{max} \}_{o=1}^M$ to partition a finite range of the observable,
\begin{equation}
\mathcal{O}_0^\mathrm{min} < \mathcal{O}_0^\mathrm{max} = \mathcal{O}_1^\mathrm{min} < \ldots < \mathcal{O}_{M-1}^\mathrm{max} = \mathcal{O}_M^\mathrm{min} < \mathcal{O}_M^\mathrm{max} \text{.}
\label{eq:bins-of-diff-xsection}
\end{equation}
The Ellis--Sexton scale $Q^2$, if chosen dynamically, depends on the phase space, however we assume the fractions $\xi_\mathrm{R}$ and $\xi_\mathrm{F}$ to be  phase-space constants in any case.
This allows for variations around the central scale choice $\xi_\mathrm{R} = \xi_\mathrm{F} = 1$, but it does not otherwise allow for arbitrary changes of the scale.
The terms with powers $m > 0$ and $n > 0$ vanish for the central scale choice and are only required for variations of the factorisation and renormalisation scales.
To estimate the perturbative QCD uncertainty --- no EW uncertainty is covered by this method --- one typically uses scale variations. Common prescriptions are
7-point and 9-points scale variations, which evaluate the cross section using
respectively the following values
  \begin{align}
    (\xi_\mathrm{R}, \xi_\mathrm{F})_{\mathrm{7-pt}}
    & \in
    \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr) \right\} \text{;}
    \label{eq:7pt}\\
    (\xi_\mathrm{R}, \xi_\mathrm{F})_{\mathrm{9-pt}}
    & \in
    \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr), \bigl( \tfrac{1}{2}, 2 \bigr), \bigl(2, \tfrac{1}{2} \bigr) \right\}  \text{.}
    \label{eq:9pt}
   \end{align}
The (asymmetric) uncertainties are then given as the minimum and maximum value (the envelope), measured from the central value $(1, 1)$.
As is clear from eq.~\eqref{eq:expansion}, the EW coupling $\alpha$ is assumed not to be a dynamically varying coupling, but instead a constant over phase space.
This, however, includes the most common choices of the coupling, which are (not necessarily in this order), $\alpha (0)$, $\alpha (M_\mathrm{Z})$, and $\alpha_{G_\mu}$.

The problem that \textsc{PineAPPL} solves can now be described: approximately reconstruct the functions
\begin{equation}
w_{ab}^{(k,l,m,n,o)} \left( x_1, x_2, Q^2 \right)
\label{eq:weight-map}
\end{equation}
of eq.~\eqref{eq:differential-cross-section} from a set of $N$ function evaluations for specific momentum fractions, scales, and values of the observable
\begin{equation}
\left\{ x_1^{(i)}, x_2^{(i)}, Q^2_i, \mathcal{O}_i \right\}_{i=1}^N \text{,} \label{eq:phase-space-weights}
\end{equation}
given by the MC integrator together with the corresponding value of the weights, eq.~\eqref{eq:weight-map}.
This problem is solved by finding an appropriate representation of eq.~\eqref{eq:phase-space-weights} and is described in section~\ref{sec:grid-representation}.
Using eq.~\eqref{eq:expansion} and
\begin{multline}
\frac{\mathrm{d} \sigma}{\mathrm{d} \mathcal{O}} (\mathcal{O}, \xi_\mathrm{R}, \xi_\mathrm{F}) \\
= \sum_{a,b} \int_0^1 \mathrm{d} x_1 \int_0^1 \mathrm{d} x_2 \int_{Q^2_\mathrm{min}}^{Q^2_\mathrm{max}} \mathrm{d} Q^2 \, f_a (x_1, \xi_\mathrm{F}^2 Q^2) f_b (x_2, \xi_\mathrm{F}^2 Q^2) \sigma_{ab} (x_1, x_2, Q^2, \xi_\mathrm{R}, \xi_\mathrm{F}) \text{,}
\label{eq:pineappl-convolution}
\end{multline}
\textsc{PineAPPL} can then quickly calculate hadronic cross sections for an arbitrary number of PDF sets and perform scale variations.
